Policing for Profit and the Undermining of Democracy\thanks{Prepared in preparation for the 2021 Annual Meeting of the American Sociological Association. Please do not circulate without permission.}

### 2021-01-28 markdown/fees_fines_to.Rmd
post "concentrated criminalization" change

  Despite this increase in research, relatively little work has explored the effects of one of the most routine interactions Americans have with police: namely, ticketing for low-level offenses. In the aftermath of the 2008 financial crisis, cash-strapped municipalities increasingly turned to fees and fines collections to fund local services. The democratic effects of this change, however, have gone largely unstudied. Using data from the Census of Governments and the national voter file, we fill this hole in the literature. Our preliminary analyses indicate that municipalities with higher fees and fines collections have lower turnout, and that this relationship is likely causal. Moreover, we find that municipalities with higher fees and fines have a *lower* racial turnout gap. Taken together, our project raises important questions about the (racial) representational effects of policing for profit.
---
\pagenumbering{gobble}
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(kevostools)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(ggpubr)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
### Introduction {-}

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are also *indirectly* disenfranchised due to over-incarceration and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities [@Lee2014; @Weaver2010; @Morris2020]. This relationship is complicated, however. In New York City, concentrated stop-and-frisk activity was associated with *reduced* turnout in Congressional elections, but *increased* turnout in the 2008 presidential election as well as the local mayoral election in 2013 [@Laniyonu2019]. It may also be the case that proximal contact differs from personal contact with the criminal legal system --- whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters --- but her survey questions specifically excluded “minor traffic stops,” a primary site of police ticketing [@Walker2014].

>>>>

  Despite this increase in research, relatively little work has explored the effects of one of the most routine interactions Americans have with police: namely, ticketing for low-level offenses. In the aftermath of the 2008 financial crisis, cash-strapped municipalities increasingly turned to fees and fines collections to fund local services. The democratic effects of this change, however, have gone largely unstudied. Using data from the Census of Governments and the national voter file, we fill this hole in the literature. Our preliminary analyses indicate that municipalities with higher per-capita fees and fines collections have lower turnout, and that this relationship is likely causal. Moreover, we find that municipalities with higher fees and fines have a *lower* racial turnout gap. Taken together, our project raises important questions about the (racial) representational effects of policing for profit.
---
\pagenumbering{gobble}
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(kevostools)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(ggpubr)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
### Introduction {-}

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are also *indirectly* disenfranchised due to concentrated criminalization and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities [@Lee2014; @Weaver2010; @Morris2020]. This relationship is complicated, however. In New York City, concentrated stop-and-frisk activity was associated with *reduced* turnout in Congressional elections, but *increased* turnout in the 2008 presidential election as well as the local mayoral election in 2013 [@Laniyonu2019]. It may also be the case that proximal contact differs from personal contact with the criminal legal system --- whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters --- but her survey questions specifically excluded “minor traffic stops,” a primary site of police ticketing [@Walker2014].

### 2021-01-28 markdown/fees_fines_to.tex
post "concentrated criminalization" change

Despite this increase in research, relatively little work has explored the effects of one of the most routine interactions Americans have with police: namely, ticketing for low-level offenses. In the aftermath of the 2008 financial crisis, cash-strapped municipalities increasingly turned to fees and fines collections to fund local services. The democratic effects of this change, however, have gone largely unstudied. Using data from the Census of Governments and the national voter file, we fill this hole in the literature. Our preliminary analyses indicate that municipalities with higher fees and fines collections have lower turnout, and that this relationship is likely causal. Moreover, we find that municipalities with higher fees and fines have a \emph{lower} racial turnout gap. Taken together, our project raises important questions about the (racial) representational effects of policing for profit.
\end{abstract}

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are also \emph{indirectly} disenfranchised due to over-incarceration and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities (Lee, Porter, and Comfort \protect\hyperlink{ref-Lee2014}{2014}; Weaver and Lerman \protect\hyperlink{ref-Weaver2010}{2010}; Morris \protect\hyperlink{ref-Morris2020}{2020}). This relationship is complicated, however. In New York City, concentrated stop-and-frisk activity was associated with \emph{reduced} turnout in Congressional elections, but \emph{increased} turnout in the 2008 presidential election as well as the local mayoral election in 2013 (Laniyonu \protect\hyperlink{ref-Laniyonu2019}{2019}). It may also be the case that proximal contact differs from personal contact with the criminal legal system --- whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters --- but her survey questions specifically excluded ``minor traffic stops,'' a primary site of police ticketing (Walker \protect\hyperlink{ref-Walker2014}{2014}).

>>>>

Despite this increase in research, relatively little work has explored the effects of one of the most routine interactions Americans have with police: namely, ticketing for low-level offenses. In the aftermath of the 2008 financial crisis, cash-strapped municipalities increasingly turned to fees and fines collections to fund local services. The democratic effects of this change, however, have gone largely unstudied. Using data from the Census of Governments and the national voter file, we fill this hole in the literature. Our preliminary analyses indicate that municipalities with higher per-capita fees and fines collections have lower turnout, and that this relationship is likely causal. Moreover, we find that municipalities with higher fees and fines have a \emph{lower} racial turnout gap. Taken together, our project raises important questions about the (racial) representational effects of policing for profit.
\end{abstract}

@@

Recent work from scholars in sociology and political science, however, indicates that many more voters are also \emph{indirectly} disenfranchised due to concentrated criminalization and the politically chilling effects of the carceral state. Policing and incarceration shape political socialization: personal contact with police or incarceration is associated with a reduced likelihood of political participation, and that alienating effect extends even to people who have incarcerated family members or who simply reside in affected communities (Lee, Porter, and Comfort \protect\hyperlink{ref-Lee2014}{2014}; Weaver and Lerman \protect\hyperlink{ref-Weaver2010}{2010}; Morris \protect\hyperlink{ref-Morris2020}{2020}). This relationship is complicated, however. In New York City, concentrated stop-and-frisk activity was associated with \emph{reduced} turnout in Congressional elections, but \emph{increased} turnout in the 2008 presidential election as well as the local mayoral election in 2013 (Laniyonu \protect\hyperlink{ref-Laniyonu2019}{2019}). It may also be the case that proximal contact differs from personal contact with the criminal legal system --- whereas direct experience with criminalization (and incarceration in particular) reduces the likelihood of voting, proximal or indirect experience produces more mixed results. Hannah Walker suggests that police contact that does not lead to a criminal conviction may even mobilize voters --- but her survey questions specifically excluded ``minor traffic stops,'' a primary site of police ticketing (Walker \protect\hyperlink{ref-Walker2014}{2014}).
